Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1]}

### 2017-12-15 paper/paper.tex
More things

\begin{abstract}
    Modern subgraph isomorphism solvers use value-ordering heuristics to direct a constraint
    programming style search. Although these heuristics are far from perfect (particularly early on
    in search), discrepancy-based searches are too expensive to be effective in this setting.
    However, we show that a random value-ordering heuristic with restarts beats conventional search
    with the best-known tailored heuristic. This motivates the introduction of a slightly-random
    value-ordering heuristic with similar behaviour to the tailored heuristic: when combined with
    restarts, this new search strategy is over ?? a hundred times more effective on satisfiable
    instances, and does not significantly decrease performance on unsatisfiable instances. Finally,
    we show that this strategy also improves two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is far from perfect---and even if degree were a perfect source of information,
many target graphs do not have a large degree distribution, so we would still need to determine
\emph{which} vertex of highest degree would be best.

We could try to introduce tiebreaking mechanisms to further refine this heuristic. However, given
the extreme cost of making an incorrect branching choice early on in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of the changes we propose are shown in
\cref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering with restarts
(\cref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing clear benefits in two maximum common induced subgraph algorithms.

@@

    \nl \While{$\KwSty{true}$}{

@@

        \nl \tikzmark{restart1s}$\KwSty{global}~(R, R') \gets (0, \textnormal{the next restart sequence value})$\tikzmark{restart1e} \;

@@

    \nl \tikzmark{restart6s}\lIf{$(R \gets R + 1) = R'$}{$\KwSty{return}~\KwSty{restart}\tikzmark{restart6e}$}

@@

        \nl \tikzmark{restart7s}trigger nogoods with watch $(v \mapsto v')$\tikzmark{restart7e} \;

@@

\subsection{Ordering Heuristics}

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.  For value ordering, which occurs on
\siplineref{valueordering} of \cref{algorithm:sip}, we will compare four strategies:

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: we expect degree to beat random, and random to beat anti.

@@

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families \citep{DBLP:journals/cviu/DamiandSHJS11} because some earlier solvers find many
of these instances extremely hard. We would like to have a larger number of satisfiable instances in
our test set, and so we include all pattern / target pairs. This gives a total of 14,621 instances
(rather than the original 5,725). At least ?? of these instances are known to be satisfiable, and at
least ?? are unsatisfiable.

\subsection{Our Solver is Good}

@@

?? Quick intro to discrepancy search
\citep{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07,DBLP:journals/jea/ProsserU11}.
Discrepancy search is about value ordering, and it mostly looks at optimisation problems and at
problems where an optimality proof is not needed.

?? Talk about how we're not dealing with a binary search tree, so we could either make it binary, or
count one discrepancy each time the degree changes, or count every ``against'' branch as being just
one discrepancy regardless.

?? \cref{figure:scatter-dds}. Sometimes extremely beneficial on satisfiable instances. However, on both
unsatisfiable and satisfiable instances, the overheads are often large, and the algorithm is not
better in aggregate (even when only considering satisfiable instances).

\subsection{Value Ordering Heuristics}

@@

In \cref{figure:value-ordering-heuristics} we show the cumulative number of instances solved over
time using the four different value-ordering heuristics. As expected, the degree heuristic is much
better (?? give aggregate speedups for both all and just satisfiable instances) than random
ordering, which is in turn ?? times better than the anti-heuristic.

The performance of the two degree heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Cref{figure:scatter-heuristics} gives a more detailed look at these results. In this plot,
satisfiable and unsatisfiable instances are shown using coloured shapes and black dots respectively.
On the $x$-axis we measure the degree heuristic, whilst on the $y$-axis we measure the biased
heuristic. Points on the outer axes represent instances which timed out with one heuristic but not
the other.  This plot shows that despite the aggregate performance being very similar, on an case by
case basis, the two heuristics can make a large difference to the performance of satisfiable
instances.

\subsection{Restarts}

?? Quick intro to restarts, mentioning how it's important to change something when restarting.
Interestingly usually that thing is variable-ordering heuristics, not value-ordering
\citep{DBLP:conf/ijcai/LecoutreSTV07,DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
?? Does anyone at all do this on value ordering explicitly? And if not, why not? There's
\citet{DBLP:conf/flairs/RazgonOP07} at least. ?? There's a connection here to confidence-based work
stealing \citep{DBLP:conf/cp/ChuSS09} too. But what? They talk about confidence in a heuristic, but
not that different values could be ranked differently.

?? We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93}, because everyone else does. We count
the number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop) to
decide when to restart. Following established wisdom, we multiply each item in the Luby sequence by
a magic constant. Preliminary experiments demonstrated that this is useful, but we failed to
determine a principled way of selecting the constant's value, so we fell back on divine revelation
and set it to 666. As with everyone else, we are not entirely clear why we are doing this: it could
be to allow the solver to spend more time deep in search, it could be to reduce the overheads from
repeatedly visiting inner nodes, or it could be to keep the nogoods small.

?? We just use two watched literals \citep{DBLP:conf/dac/MoskewiczMZZM01}. Could use increasing
nogoods \citep{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17} etc, but we are not confident enough
in our programming abilities to implement these correctly, and we are not maintaining arc
consistency anyway. We also really do not want to propagate on negative decisions, because we do not
want to iterate over every removed value.

?? Talk a bit about how it fits into \cref{algorithm:sip}.

\subsection{Restarts Make it Better}

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In
other words, although having a good variable-ordering heuristic is beneficial, introducing
randomness into the search is better, if it is done alongside a mechanism to avoid heavy commitment
to any particular random choice. \Cref{figure:scatter-random} looks at this in more detail.  The
performance on unsatisfiable instances is the same. However, although better in aggregate, we see
many satisfiable instances performing substantially worse. ?? Quantify this a bit.

@@

\subsection{Further Experiments}

\begin{itemize}
    \item The choice of Luby multiplier is not hugely important, although 100 is too small and 1,000
        is too big.
    \item The base of the exponential can be $2$ or $e$ or $3$ or $1.5$ and it makes little
        difference. Using a polynomial weighting is bad, and using a base of $|D_v|$ is bad.
    \item Using geometric restarts instead of Luby does not give as much of a benefit. A constant
        restart policy is also fine (and terminates on unsatisfiable instances because we count
        backtracks).
    \item Using rank with restarts is better than random with restarts, but not as good as using
        biased with restarts.
    \item Nogood recording is important.
    \item Nogood recording is cheap.
    \item Discarding subsumed nogoods may be done cheaply, if memory space is a concern.
\end{itemize}

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{McSplit speedups ?? partial data}\label{figure:mcsplit}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \caption{k${\downarrow}$ speedups ?? missing data}\label{figure:kdown}
\end{figure}

@@

\clearpage
\bibliographystyle{named}
\bibliography{dblp}

>>>>

\begin{abstract}
    Modern subgraph isomorphism solvers use value-ordering heuristics to direct a constraint
    programming style search. Although these heuristics are far from perfect, discrepancy-based
    searches are too expensive to be effective in this setting.  However, we show that a random
    value-ordering heuristic with restarts beats conventional search with the best-known tailored
    heuristic. This motivates the introduction of a slightly-random value-ordering heuristic with
    similar behaviour to the tailored heuristic: when combined with restarts, this new search
    strategy is over ?? times more effective on satisfiable instances, and does not
    significantly decrease performance on unsatisfiable instances. Finally, we show that this
    strategy also improves two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early on in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of the changes we propose are shown in
\cref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering with restarts
(\cref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing clear benefits in two maximum common induced subgraph algorithms.

@@

    \nl \While{$\KwSty{true}$\siplabel{restartsloop}}{

@@

        \nl \tikzmark{restart1s}$\KwSty{global}~(R, R') \gets (0, \textnormal{the next Luby sequence value})$\tikzmark{restart1e} \;

@@

    \nl \tikzmark{restart6s}\lIf{$(R \gets R + 1) = R'$}{$\KwSty{return}~\KwSty{restart}\tikzmark{restart6e}$\siplabel{countrestarts}}

@@

        \nl \tikzmark{restart7s}trigger nogoods with watch $(v \mapsto v')$\tikzmark{restart7e}\siplabel{2wl} \;

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.  For value ordering, which occurs on
\siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from highest
degree to lowest degree; we return to value ordering in the second half of this paper.

?? note that variable ordering affects unsat, whereas value ordering affects only sat.

@@

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least ?? of these instances are known to be satisfiable, and at least ?? are unsatisfiable.

\subsection{Baseline Performance}

@@

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\citep{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97} are alternatives to
backtracking search that initially search disallowing all discrepancies, and then retry search
allowing an increasing number of discrepancies at each iteration until either a solution is found or
unsatisfiability is proven. These schemes assume that value-ordering heuristics are usually
reliable, and that most solutions can be found with only a small number of discrepancies; in such
cases, the heavy commitment to early branching choices made by conventional backtracking search can
be extremely costly.

\Cref{figure:scatter-dds} shows the effects of adding \citet{DBLP:conf/ijcai/Walsh97}'s
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

The points well below the diagonal line and along the right-hand axis show that DDS can sometimes be
extremely beneficial on satisfiable instances.  However, on both unsatisfiable and most satisfiable
instances, the overheads can be extremely large, and DDS is much worse in aggregate (even when only
considering satisfiable instances).

These large overheads are to be expected: discrepancy searches are aimed primarily at getting better
feasible solutions in optimisation problems which are too large for a proof of optimality to be a
realistic prospect, and they are not well-suited for unsatisfiable decision problems. However, the
extremely large gains on some satisfiable instances confirm our suspicions that we should look at
value-ordering heuristics in more detail.

\subsection{Value-Ordering Heuristics}

@@

In \cref{figure:value-ordering-heuristics} we show the cumulative number of satisfiable instances
solved over time using four different value-ordering heuristics:

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: as expected, degree beats random, and random beats anti.

The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Cref{figure:scatter-heuristics} gives a more detailed look at these results. This scatter plot
shows that despite the aggregate performance being very similar, on an case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

\subsection{Adding Restarts to the Algorithm}

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and keep the
variable ordering fixed.\footnote{It may be possible to further improve \cref{algorithm:sip} by also
introducing randomness or some form of learning into its variable-ordering heuristic. However,
introducing a second change would considerably complicate the empirical analysis. We therefore leave
this change as future work.} Using restarts on value-ordering heuristics appears to be uncommon,
although \citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from
restarts, and \citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel search.

The shaded code shows how to introduce restarts to \cref{algorithm:sip}. The recursive
$\FuncSty{search}$ procedure is modified to return either true (if the instance is satisfiable),
false (unsatisfiable), or a third value if some threshold was reached without the solution being
known.  The main function then calls $\FuncSty{search}$ in a loop (\siplineref{restartsloop}), until
a solution is found.

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant.
Preliminary experiments demonstrated that a constant 100 is slightly too small and 1,000 is slightly
too large, but lacking a more principled way of selecting the constant's value, we fell back on
divine revelation and use a multiplier of 666. (We also tried a geometric scheme. The results were
less favourable---it appears that value-ordering heuristics benefit from a very aggressive restart
strategy.)

The remainder of the changes are to introduce nogoods. To avoid exploring portions of the search
space that we have already visitied, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees. We generate simple decision nogoods, as follows. The $B$
argument to $\FuncSty{search}$ tracks the branching choices made so far, in the form of a CNF clause
with each literal being a guessed $(v \mapsto v')$ assignment. Upon backtracking due to a decision to
restart, we post a nogood of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x')
\Rightarrow \bot$ for the current branch, and for every branch to its left in the search tree.

Finally, we use the two watched literals technique \citep{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods (\siplineref{2wl}). This has two benefits: the propagation complexity does
not particularly depend upon the number of stored nogoods, and it does not require any work upon
backtracking.  Other more sophisticated nogood generation and propagation schemes exist
\citep{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}, but it is not clear whether these will be
beneficial in a setting where we are not maintaining arc consistency.

\subsection{The Effects of Restarts}

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good variable-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts. ?? Quantify this a bit

@@

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{McSplit speedups ?? partial data, can cut for space and just use
    cumulatives if necessary}\label{figure:mcsplit}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \caption{k${\downarrow}$ speedups ?? missing data, can cut for space and just use
    cumulatives if necessary}\label{figure:kdown}
\end{figure}

@@

\bibliographystyle{named}
\bibliography{dblp}
